---
title: "COVID-19 Rapid Antigen Tests and Bayes' Theorem - are you positive that you're certain that you're positive?"
author: Tom
date: '2022-07-28'
slug: covid19-rapid-antigen-tests-and-bayes-theorem
categories:
  - Fluff
tags:
  - COVID-19
  - Rapid Antigen Test
  - Bayes' Theorem
type: ''
subtitle: ''
image: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

A week ago I went to dinner with a group of friends.
The next day the person sat opposite me informed me they had tested positive for COVID-19, including a photo of their Rapid Antigen Test (RAT) cassette[^1].
A few days later I developed a dry throat, fatigue, and brain fog, but also some very concerning some flu-like symptoms... similar to my friend who had just tested positive... oh no!
Naturally I went overboard on the RATs and performed a test every day for five days.
To my relief, all five were negative and I concluded it was just a nasty cold.

[^1]: Everyone who attended was okay!
    My friend recovered well and no one else tested positive from that event.

Was I really going overboard though?
RAT tests are currently the primary means for identifying positive cases of COVID-19 in Victoria.
In the last 24 hours before the time of writing this post, [3085 cases were identified via PCR testing and 9069 probable cases via RATs](https://www.coronavirus.vic.gov.au/victorian-coronavirus-covid-19-data).
That means 3/4 positive cases were identified using RATs.

However, that gives us no information on why anyone would take five RAT tests.
What kind of maniac would do such a thing?
And who are we (i.e., you) to judge?
Well, as it turns out, we can use Bayes' Theorem to work out exactly how much of a maniac that person (i.e., I) might be!
Clearly I'm worried about the false-negative rate, and

## Bayes' Theorem

Bayes' Theorem describes the probability that an event should occur, given that a condition has (or hasn't) been met.

$Pr(A|B) = \frac{Pr(B|A)Pr(A)}{Pr(B)}$

$Pr(B)$ is found with the following equation:

$Pr(B) = Pr(B|A)Pr(A) + Pr(D|\bar{A})Pr(\bar{A})$

It applies well to medical screening and defective item testing - the two things I want to know more about.
There's a few things that were found in Leiner et al. (2021): The specificity (i.e., the True-Positive rate) was 62.3%, and the specificity (i.e., the True-Negative) was 97.6%.
If we plug these numbers into the formula above, we get the probability that a person is infected with COVID-19 after testing positive on a RAT: \n

```{r}
# covid bayesian test:

vic_population <- 6740000 # 
vic_infected <- 69057 # from https://www.coronavirus.vic.gov.au/victorian-coronavirus-covid-19-data

p_infected <- vic_infected/vic_population
p_positive_infected <- 0.623  # Sensitivity / True Positive
p_negative_safe <- 0.976      # Specificity / True Negative
p_positive_safe <- (1-p_negative_safe)  # False Negative

p_positive <- (p_positive_infected*p_infected) + (p_positive_safe*(1-p_infected))
p_infected_positive <- (p_positive_infected*p_infected)/p_positive

print(paste0(round(p_infected_positive * 100, 2), "%"))
```

\n

So, the posterior probability that I have contracted COVID-19 after having tested positive is only 21.18%.
That's... extremely shocking.
This suggests RAT tests are incorrect about 4 out of 5 times!
So if I perform 5 RATs, I should expect exactly one of them to have a true positive result, yes?
And that means my method is not stupid and I'm not a maniac, YES??
\n

Sadly, no.

```{r}
# covid bayesian test:

vic_population <- 6740000 # 
vic_infected <- 69057 # from https://www.coronavirus.vic.gov.au/victorian-coronavirus-covid-19-data

p_infected <- vic_infected/vic_population
p_positive_infected <- 0.623  # Sensitivity / True Positive
p_negative_safe <- 0.976      # Specificity / True Negative
p_positive_safe <- (1-p_negative_safe)  # False Negative

p_positive <- (p_positive_infected*p_infected) + (p_positive_safe*(1-p_infected))
p_infected_positive <- (p_positive_infected*p_infected)/p_positive

###

p_safe <- (1-p_infected)
p_negative_infected <- (1-p_positive_infected)  # Sensitivity / True Positive

p_negative <- (p_negative_infected*p_infected) + (p_negative_safe*(1-p_infected))
p_infected_negative <- (p_negative_infected*p_infected)/p_negative


print(paste0(round(p_infected_negative * 100, 2), "%"))
```

```{r}
# covid bayesian test:

vic_population <- 6740000 # 
vic_infected <- 69057 # from https://www.coronavirus.vic.gov.au/victorian-coronavirus-covid-19-data

p_infected <- vic_infected/vic_population
p_positive_infected <- 0.623  # Sensitivity / True Positive
p_negative_safe <- 0.976      # Specificity / True Negative
p_positive_safe <- (1-p_negative_safe)  # False Negative

p_positive <- (p_positive_infected*p_infected) + (p_positive_safe*(1-p_infected))
p_infected_positive <- (p_positive_infected*p_infected)/p_positive

###

p_safe <- (1-p_infected)
p_negative_infected <- (1-p_positive_infected)  # Sensitivity / True Positive

p_negative <- (p_negative_infected*p_infected) + (p_negative_safe*(1-p_infected))
p_safe_negative <- (p_negative_safe*p_safe)/p_negative


print(paste0(round(p_safe_negative * 100, 2), "%"))
```

```{r}
# covid bayesian test:

vic_population <- 6740000 # 
vic_infected <- 69057 # from https://www.coronavirus.vic.gov.au/victorian-coronavirus-covid-19-data

p_infected <- vic_infected/vic_population
p_positive_infected <- 0.623  # Sensitivity / True Positive
p_positive_infected <- 0.623  # Sensitivity / True Positive
p_negative_safe <- 0.976      # Specificity / True Negative
p_positive_safe <- (1-p_negative_safe)  # False Negative

p_positive <- (p_positive_infected*p_infected) + (p_positive_safe*(1-p_infected))
p_infected_positive <- (p_positive_infected*p_infected)/p_positive

print(paste0(round(p_infected_positive * 100, 2), "%"))
```

#### False Negatives:

#### False Positives:
